Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-01-15 paper/paper.tex
More writings

\usepackage{cleveref}

@@

@@

?? loops, be really careful with this

\paragraph{Constraint programming.} ?? A constraint programming algorithm.

\paragraph{Degree-based invariants for homomorphisms.} The use of constraint programming techniques
for finding homomorphisms relies upon exploiting various invariants. It is straightforward to verify the following.

\begin{proposition}\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from a $G$ to  $H$. Let
    $v$ be any vertex. Then:
    \begin{enumerate}
        \item $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
        \item Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own
           .
    \end{enumerate}
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these also hold for subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}.
Every recent
constraint programming solver makes use of the of the above invariants for subgraph isomorphisms.
However, these invariants do \emph{not} hold for homomorphisms in general. As a simple counter-example, a
homomorphism may map a star graph (one central vertex of high degree, which is adjacent to many
other vertices, each of which has no other neighbours) onto a single edge.

The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a pattern vertex, any value representing target vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.

@@

\begin{proposition}For the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph
    $H$, the following constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at least the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two. Local injectivity is trickier. The second property does not hold in full
generality, because, for example, a path of length three may be mapped onto a triangle whilst
respecting local injectivity. However, a weaker result holds:

\begin{proposition}For the problem of finding a locally injective graph homomorphism $i$ from a
    graph $G$ to a graph $H$, for any pair of vertices $v, w \in \vertexset(G)$, if there are at
    least $k$ simple paths of length exactly two between $v$ and $w$, then there must be at least
    $k$ simple paths of length exactly two between $i(v)$ and $i(w)$.
\end{proposition}

@@

?? Corollary: $G^d$ graphs.

@@

\begin{proposition}[Cliques are preserved]\label{proposition:clique}
    Let $i$ be a homomorphism from $G$ to $H$ where $H$ does
    not contain any loops. Let $S\subset V$ be such that $G[S]=K_k$. Then $H[i(S)]=K_k$. 
\end{proposition}

@@

>>>>

\usepackage{cleveref}
\crefname{figure}{figure}{figures}
\Crefname{figure}{Figure}{Figures}

\usepackage{tikz}
\usetikzlibrary{shapes}

@@

\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}

@@

\paragraph{Loops.} A vertex which is adjacent to itself is called a \emph{loop}. By a careful reading
of the definition, any homomorphism must map loops onto loops---and indeed, the Glasgow Subgraph
Solver and this paper take this approach (although some other algorithm implementations do not). It
therefore follows that for the homomorphism \emph{decision} problem where there are no injectivity
constraints, any problem instance which has a loop in the codomain graph is trivially satisfiable,
although the \emph{counting} problem remains \#\P-hard \cite{DBLP:journals/rsa/DyerG00}.

\paragraph{Constraint programming.} A \emph{constraint satisfaction problem} is defined in terms of
a set of \emph{variables}, each of which has a \emph{domain} of possible \emph{values}, together
with a set of constraints; the goal is to give each variable a value from its domain, whilst
satisfying all of the constraints. Homomorphism problems have a natural representation as a
constraint satisfaction problem: we have a variable for each vertex in the domain graph, whose
values range over the codomain graph, and a set of constraints saying that adjacent vertices must be
mapped to adjacent vertices. For the injective variants, one or more \emph{all-different}
constraints are also present. A typical constraint programming approach to solving such a problem is
to combine inference through constraint propagation with an intelligent backtracking search.

\paragraph{Degree-based invariants for homomorphisms.} Constraint programming techniques for finding
homomorphisms can be made much more effective by exploiting certain invariants. It is
straightforward to verify the following.

\begin{proposition}\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from a $G$ to  $H$. Let
    $v$ be any vertex. Then:
    \begin{enumerate}
        \item $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
        \item Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own.
    \end{enumerate}
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these invariants also hold for
subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}; indeed, every recent constraint
programming approach for subgraph isomorphism finding make use of degrees and neighbourhood degree
sequences. The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a pattern vertex, any value representing target vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.

\begin{figure}
    \hfill\,\begin{minipage}{3cm}
    \begin{tikzpicture}
        \node [draw, fill, circle, inner sep=1.5pt] (A1) at (0, 0) {};
        \node [draw, circle, inner sep=1.5pt] (A2) at (0:0.3) {};
        \node [draw, circle, inner sep=1.5pt] (A3) at (60:0.3) {};
        \node [draw, circle, inner sep=1.5pt] (A4) at (120:0.3) {};
        \node [draw, circle, inner sep=1.5pt] (A5) at (180:0.3) {};
        \node [draw, circle, inner sep=1.5pt] (A6) at (240:0.3) {};
        \node [draw, circle, inner sep=1.5pt] (A7) at (300:0.3) {};

        \draw (A1) -- (A2);
        \draw (A1) -- (A3);
        \draw (A1) -- (A4);
        \draw (A1) -- (A5);
        \draw (A1) -- (A6);
        \draw (A1) -- (A7);

        \draw [->] (0.7, 0) -- (1.1, 0);

        \draw (1.5, 0) node [draw, fill, circle, inner sep=1.5pt] (B1) at ++ (90:0.15) {};
        \draw (1.5, 0) node [draw, circle, inner sep=1.5pt] (B2) at ++ (270:0.15) {};
        \draw (B1) -- (B2);
    \end{tikzpicture}\end{minipage}\hfill\begin{minipage}{3cm}
    \begin{tikzpicture}
        \node [draw, fill, circle, inner sep=1.5pt] (A1) at (0, 0) {};
        \node [draw, inner sep=1.5pt] (A2) at (0.3, 0) {};
        \node [draw, inner sep=1.5pt] (A3) at (0.3, -0.2) {};
        \node [draw, inner sep=1.5pt] (A4) at (0.3, 0.2) {};
        \node [draw, circle, inner sep=1.5pt] (A5) at (0.6, 0) {};

        \draw (A1) -- (A2);
        \draw (A1) -- (A3);
        \draw (A1) -- (A4);
        \draw (A2) -- (A5);
        \draw (A3) -- (A5);
        \draw (A4) -- (A5);

        \draw [->] (1.0, 0) -- (1.4, 0);

        \node [draw, fill, circle, inner sep=1.5pt] (B1) at (1.8, 0) {};
        \node [draw, inner sep=1.5pt] (B2) at (2.1, 0) {};
        \node [draw, circle, inner sep=1.5pt] (B5) at (2.4, 0) {};

        \draw (B1) -- (B2);
        \draw (B2) -- (B5);
    \end{tikzpicture}\end{minipage}\hfill\,

    \bigskip

    \,\hfill\begin{minipage}{3cm}
    \begin{tikzpicture}
        \node [draw, fill, circle, inner sep=1.5pt] (A1) at (0, 0) {};
        \node [draw, circle, inner sep=1.5pt] (A2) at (0.3, 0) {};
        \node [draw, inner sep=1.5pt] (A3) at (0.6, 0) {};
        \node [draw, fill, circle, inner sep=1.5pt] (A4) at (0.9, 0) {};

        \draw (A1) -- (A2);
        \draw (A2) -- (A3);
        \draw (A3) -- (A4);

        \draw [->] (1.3, 0) -- (1.7, 0);

        \draw (2.2, 0) node [draw, fill, circle, inner sep=1.5pt] (B1) at ++ (120:0.3) {};
        \draw (2.2, 0) node [draw, circle, inner sep=1.5pt] (B2) at ++ (240:0.3) {};
        \draw (2.2, 0) node [draw, inner sep=1.5pt] (B3) at ++ (0:0.3) {};

        \draw (B1) -- (B2);
        \draw (B2) -- (B3);
        \draw (B3) -- (B1);
    \end{tikzpicture}\end{minipage}\hfill\begin{minipage}{3cm}
        \begin{tikzpicture}
        \node [draw, fill, circle, inner sep=1.5pt] (A1) at (0, 0) {};
        \node [draw, inner sep=1.5pt] (A2) at (0.3, -0.2) {};
        \node [draw, fill, inner sep=1.5pt] (A3) at (0.3, 0.2) {};
        \node [draw, circle, inner sep=1.5pt] (A4) at (0.6, 0) {};
        \node [draw, diamond, fill, inner sep=1.5pt] (A5) at (0.9, -0.2) {};
        \node [draw, diamond, inner sep=1.5pt] (A6) at (0.9, 0.2) {};
        \node [draw, fill, circle, inner sep=1.5pt] (A7) at (1.2, 0) {};

        \draw (A1) -- (A2);
        \draw (A1) -- (A3);
        \draw (A2) -- (A4);
        \draw (A3) -- (A4);
        \draw (A4) -- (A5);
        \draw (A4) -- (A6);
        \draw (A5) -- (A7);
        \draw (A6) -- (A7);

        \draw [->] (1.4, 0) -- (1.7, 0);

        \node [draw, fill, circle, inner sep=1.5pt] (B1) at (2, 0) {};
        \node [draw, fill, inner sep=1.5pt] (B2) at (2.3, -0.15) {};
        \node [draw, diamond, inner sep=1.5pt] (B3) at (2.3, 0.15) {};
        \node [draw, diamond, fill, inner sep=1.5pt] (B4) at (2.3, -0.45) {};
        \node [draw, inner sep=1.5pt] (B5) at (2.3, 0.45) {};
        \node [draw, circle, inner sep=1.5pt] (B6) at (2.6, 0) {};

        \draw (B1) -- (B2);
        \draw (B1) -- (B3);
        \draw (B1) -- (B4);
        \draw (B1) -- (B5);
        \draw (B6) -- (B2);
        \draw (B6) -- (B3);
        \draw (B6) -- (B4);
        \draw (B6) -- (B5);
    \end{tikzpicture}\end{minipage}\hfill\,

    \caption{Counter-examples for various properties which are not invariants. The different shapes
        for vertices show a mapping, with vertices being mapped to vertices of the same shape. The
        top left example shows that degree is not preserved in a homomorphism; the top right that path
        counts are not preserved in a homomorphism; the bottom left that path counts of length three
        are not preserved in a locally injective homomorphism; and the bottom right that a locally
        injective homomorphism $i$ can give rise to a homomorphism $i^{2,2}$ which is not locally
        injective.} \label{figure:counterexamples}
\end{figure}

However, these invariants do \emph{not} hold for homomorphisms in general. As a simple
counter-example, a homomorphism may map a star graph onto a single edge, as in
\cref{figure:counterexamples} (top left).

@@

\begin{proposition}\label{proposition:distances}For the problem of finding a subgraph isomorphism
    $i$ from a graph $G$ to a graph $H$, the following constraints are implied for any pair of
    vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at least the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). Local injectivity is trickier.
The second property does not hold in full generality, because, for example, a path of length three
may be mapped onto a triangle whilst respecting local injectivity (\cref{figure:counterexamples},
bottom left).  However, we observe that a weaker result holds:

\begin{proposition}\label{proposition:paths}For the problem of finding a locally injective graph
    homomorphism $i$ from a graph $G$ to a graph $H$, for any pair of vertices $v, w \in
    \vertexset(G)$, if there are at least $k$ simple paths of length exactly two between $v$ and
    $w$, then there must be at least $k$ simple paths of length exactly two between $i(v)$ and
    $i(w)$.
\end{proposition}

@@

Rather than using distance properties directly, \citet{DBLP:conf/cp/McCreeshP15} introduced
the notion of \emph{supplemental} graphs, with the idea that a constraint programming algorithm can
search for a mapping which is simultaneously a subgraph isomorphism between several different pairs
of graphs. Let $G^d$ be the graph with the same set of vertices as $G$, but an edge between vertices
$v$ and $w$ if the distance between $v$ and $w$ in $G$ is at most $d$.  Similarly, let $G^{n,\ell}$
be the graph with the same set of vertices as $G$, but an edge between vertices $v$ and $w$ if there
are at least $n$ simple paths of length exactly $\ell$ between $v$ and $w$ in $G$. Following on from
\cref{proposition:distances,proposition:paths}, we generalise the result of
\citeauthor{DBLP:conf/cp/McCreeshP15} as follows.

\begin{corollary}Any homomorphism $i$ from $G$ to $H$ gives a homomorphism $i^d$ from $G^d$ to $H^d$
    defined by $i^d(v) = i(v)$, for all $d$. Furthermore, if $i$ is locally injective, then for all
    $n$, there is a homomorphism $i^{n,2}$ from $G^{n,2}$ to
    $H^{n,2}$ defined by $i^{n,2}(v) = i(v)$.
\end{corollary}

Note carefully that $i^{n,2}$ is \emph{not} necessarily locally injective, even if $i$ is; we
illustrate a counter-example in \cref{figure:counterexamples} (bottom right). This is in contrast to
subgraph isomorphism, where $i^{n,\ell}$ is a subgraph isomorphism for all $\ell$
\cite{DBLP:conf/cp/McCreeshP15}.

@@

\begin{proposition}[Cliques are preserved]\label{proposition:clique}
    Let $i$ be a homomorphism from $G$ to $H$ where $H$ does
    not contain any loops. Let $S\subset V$ be such that $G[S]$ is a $k$-vertex clique. Then
    $H[i(S)]$ is also a $k$-vertex clique.
\end{proposition}

@@

?? Epimorphism
